# ALY 6050

# Module 3 Project - Forecasting Financial Time Series

# Thuy Nhu Thao Tran

```{r}
# Load the Required Libraries and Data
# Load necessary libraries
library(readxl)      
library(forecast)    
library(ggplot2)     
library(zoo)         

# Load the data from the Excel file
data <- read_excel("ALY6050_Module3Project_Data.xlsx", sheet = "6050_Module3Project_Data")

# Check for missing values in the dataset
colSums(is.na(data))

# Remove rows with missing values
data_clean <- na.omit(data)

# Verify that missing values are removed
colSums(is.na(data_clean))

# Extract cleaned AAPL and HON data
aapl_data_clean <- data_clean$`AAPL (Apple Inc) / $`
hon_data_clean <- data_clean$`HON (Honeywell Inc)  /  $`

```

# Part 1: Short-term Forecasting

# 1.1 Time Series Plots

```{r}
# Plot AAPL stock prices
ggplot(data_clean, aes(x = Period, y = `AAPL (Apple Inc) / $`)) +
  geom_line(color = "blue") +
  labs(title = "AAPL Stock Prices Over Time", x = "Period", y = "Price ($)")

# Plot HON stock prices
ggplot(data_clean, aes(x = Period, y = `HON (Honeywell Inc)  /  $`)) +
  geom_line(color = "red") +
  labs(title = "HON Stock Prices Over Time", x = "Period", y = "Price ($)")
```

Interpretation: The plot of AAPL stock prices and the plot of HON stock prices show an overall upward tendency with some fluctuations. AAPL exhibits higher volatility compared to HON, which appears to be more consistent. Both stocks show no obvious seasonal trends.

# 1.2 Exponential Smoothing

```{r}
# Define the exponential smoothing function
exp_smoothing <- function(data, alpha) {
  model <- HoltWinters(data, alpha = alpha, beta = FALSE, gamma = FALSE)
  forecast <- forecast(model, h = 1)
  
  # Calculate MAPE 
  mape <- mean(abs((data - forecast$fitted) / data), na.rm = TRUE) * 100
  return(list(forecast = forecast, mape = mape))
}
# Apply exponential smoothing to cleaned AAPL data 
aapl_results <- lapply(c(0.15, 0.35, 0.55, 0.75), function(alpha) exp_smoothing(aapl_data_clean, alpha))

# Apply exponential smoothing to cleaned HON data 
hon_results <- lapply(c(0.15, 0.35, 0.55, 0.75), function(alpha) exp_smoothing(hon_data_clean, alpha))

# Display MAPE results for AAPL
sapply(aapl_results, function(x) x$mape)

# Display MAPE results for HON
sapply(hon_results, function(x) x$mape)
```

Interpretation: The stock prices of AAPL and HON were subjected to exponential smoothing with α values of 0.15, 0.35, 0.55, and 0.75. In contrast to HON, which had MAPE values of 3.06, 2.35, 2.03, and 1.90, respectively, AAPL had MAPE values of 3.87, 2.47, 2.08, and 1.97. For both stocks, the most accurate predictions were obtained when α = 0.75. This variation is due to the fact that a larger α value assigns more weight to recent data, which is crucial for identifying both subtle trends in HON's stock price and quick changes in AAPL's stock price. The findings imply that current data is crucial for predicting both firms' stock values.

-\> Best MAPE for AAPL: 1.969072

-\> Best MAPE for HON: 1.895824

# 1.3 Adjusted Exponential Smoothing

```{r}
# Define the adjusted exponential smoothing function 
adj_exp_smoothing <- function(data, alpha, beta) {
  model <- HoltWinters(data, alpha = alpha, beta = beta, gamma = FALSE)
  forecast <- forecast(model, h = 1)
  
  # Calculate MAPE (handle division by zero and NA values)
  mape <- mean(abs((data - forecast$fitted) / data), na.rm = TRUE) * 100
  return(list(forecast = forecast, mape = mape))
}
# Apply adjusted exponential smoothing to cleaned AAPL data 
aapl_adj_results <- lapply(c(0.15, 0.25, 0.45, 0.85), function(beta) adj_exp_smoothing(aapl_data_clean, 0.55, beta))

# Apply adjusted exponential smoothing to cleaned HON data 
hon_adj_results <- lapply(c(0.15, 0.25, 0.45, 0.85), function(beta) adj_exp_smoothing(hon_data_clean, 0.55, beta))

# Display MAPE results for AAPL (adjusted smoothing)
sapply(aapl_adj_results, function(x) x$mape)

# Display MAPE results for HON (adjusted smoothing)
sapply(hon_adj_results, function(x) x$mape)

```

Interpretation: The stock prices of AAPL and HON were subjected to adjusted exponential smoothing with α = 0.55 and β values of 0.15, 0.25, 0.45, and 0.85. In contrast to HON, which had MAPE values of 2.06, 2.08, 2.17, and 2.28 correspondingly, AAPL had MAPE values of 2.03, 2.05, 2.17, and 2.30. For both stocks, the most accurate predictions were obtained when β = 0.15. This is due to the fact that a smaller β value eliminates noise and prevents overfitting to short-term movements, both of which are critical for capturing the general trend in steady (HON) and high-volatility (AAPL) equities. The findings imply that prediction accuracy is increased by flattening the trend component.

# Part 2: Long-term Forecasting

# 2.1 3-Period Weighted Moving Averages

```{r}
# Load necessary libraries
library(forecast)
library(ggplot2)

# Define the 3-period weighted moving average function
weighted_moving_average <- function(data, weights) {
  n <- length(data)
  forecast <- rep(NA, n)  # Initialize with NA
  weights <- weights / sum(weights)  # Normalize weights to sum to 1
  
  if (length(weights) != 3) {
    stop("Weights vector must have exactly 3 elements.")
  }
  
  for (i in 3:n) {
    forecast[i] <- sum(data[(i-2):i] * weights)
  }
  return(forecast)
}

# Ensure data is a numeric vector
aapl_data_clean <- as.numeric(aapl_data_clean)
hon_data_clean <- as.numeric(hon_data_clean)

# Define weights for the moving average
weights <- c(0.2, 0.3, 0.5)  # Weights for two periods ago, one period ago, and most recent

# Apply 3-period weighted moving average for AAPL and HON data (first 100 periods)
aapl_wma <- weighted_moving_average(aapl_data_clean[1:100], weights)
hon_wma <- weighted_moving_average(hon_data_clean[1:100], weights)

# Define training period indices for trend forecasting
train_period <- 101:252

# Linear trend forecasting for AAPL (fit using periods 101-252)
aapl_train_data <- data.frame(Period = train_period, Value = aapl_data_clean[train_period])
aapl_trend <- lm(Value ~ Period, data = aapl_train_data)

# Forecast periods 101-257
forecast_period <- 101:257
newdata_aapl <- data.frame(Period = forecast_period)
aapl_trend_forecast <- predict(aapl_trend, newdata = newdata_aapl)

# Linear trend forecasting for HON (fit using periods 101-252)
hon_train_data <- data.frame(Period = train_period, Value = hon_data_clean[train_period])
hon_trend <- lm(Value ~ Period, data = hon_train_data)

# Forecast periods 101-257
newdata_hon <- data.frame(Period = forecast_period)
hon_trend_forecast <- predict(hon_trend, newdata = newdata_hon)

# **Fix the Length Issue**
# Ensure forecasts have exactly 257 elements
aapl_forecasts <- c(aapl_wma, rep(NA, 100 - length(aapl_wma)), aapl_trend_forecast)
hon_forecasts <- c(hon_wma, rep(NA, 100 - length(hon_wma)), hon_trend_forecast)

# Extract forecasts for periods 253-257
aapl_forecasts_253_257 <- setNames(aapl_forecasts[253:257], 253:257)
hon_forecasts_253_257 <- setNames(hon_forecasts[253:257], 253:257)

# Display forecasts
cat("AAPL Forecasts for Periods 253-257:\n")
print(aapl_forecasts_253_257)

cat("HON Forecasts for Periods 253-257:\n")
print(hon_forecasts_253_257)
```

AAPL Forecasts: Because the approach significantly overstated the real prices, the AAPL estimates were not particularly accurate. This is probably because AAPL's stock price is so volatile that it is difficult to forecast using a simple weighted moving average and linear trend.

HON Forecasts: Additionally, the HON forecasts were not particularly accurate since the approach significantly underestimated the real costs. This conclusion implies that the increased trend in HON's stock price during this period may not have been well represented by the linear trend.

# 2.2 MAPEs calculation and comparison.

```{r}
# Define the linear trend forecasting function
linear_trend <- function(data, start_period) {
  n <- length(data)
  trend <- lm(data[start_period:n] ~ c(start_period:n))
  forecast <- predict(trend, newdata = data.frame(c(start_period:n)))
  return(forecast)
}

# Apply linear trend forecasting to AAPL data
aapl_trend <- linear_trend(aapl_data_clean, 101)

# Apply linear trend forecasting to HON data
hon_trend <- linear_trend(hon_data_clean, 101)

# Calculate MAPE for AAPL long-term forecast
aapl_long_mape <- mean(abs((aapl_data_clean[101:length(aapl_data_clean)] - aapl_trend) / aapl_data_clean[101:length(aapl_data_clean)])) * 100

# Calculate MAPE for HON long-term forecast
hon_long_mape <- mean(abs((hon_data_clean[101:length(hon_data_clean)] - hon_trend) / hon_data_clean[101:length(hon_data_clean)])) * 100

# Display MAPE results for long-term forecasts
aapl_long_mape
hon_long_mape

```

AAPL: The MAPE in Part 2.2 (4.778476) is much higher than the best MAPE for AAPL in Part 1 (1.969072). This finding suggests that the most accurate prediction for AAPL was produced using the exponential smoothing approach, which had an alpha of 0.75.

HON: Likewise, the MAPE in Part 2.2 (3.052453) is higher than the best MAPE for HON in Part 1 (1.895824). This result implies that the most accurate prediction for HON was likewise produced using the exponential smoothing approach, which had an alpha of 0.75.

# Part 3: Regression

# 3.1 Simple Regression

```{r}
# Perform simple regression for AAPL
aapl_regression <- lm(`AAPL (Apple Inc) / $` ~ Period, data = data_clean)
aapl_regression_forecast <- predict(aapl_regression, newdata = data.frame(Period = 1:257))

# Perform simple regression for HON
hon_regression <- lm(`HON (Honeywell Inc)  /  $` ~ Period, data = data_clean)
hon_regression_forecast <- predict(hon_regression, newdata = data.frame(Period = 1:257))

# Calculate MAPE for AAPL regression forecast
aapl_actual <- data_clean$`AAPL (Apple Inc) / $`[1:257]
aapl_regression_mape <- mean(abs((aapl_actual - aapl_regression_forecast) / aapl_actual), na.rm = TRUE) * 100

# Calculate MAPE for HON regression forecast
hon_actual <- data_clean$`HON (Honeywell Inc)  /  $`[1:257]
hon_regression_mape <- mean(abs((hon_actual - hon_regression_forecast) / hon_actual), na.rm = TRUE) * 100

# Display MAPE results for regression forecasts
aapl_regression_mape
hon_regression_mape

```

# 3.2 Residual Analysis

```{r}
# Load necessary libraries
library(stats) # For shapiro.test

# 1. Independence of Residuals
# Plot residuals vs fitted values for AAPL
aapl_residuals <- residuals(aapl_regression)
aapl_fitted <- fitted(aapl_regression)

ggplot(data = data.frame(Fitted = aapl_fitted, Residuals = aapl_residuals), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "AAPL Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals")

# Plot residuals vs fitted values for HON
hon_residuals <- residuals(hon_regression)
hon_fitted <- fitted(hon_regression)

ggplot(data = data.frame(Fitted = hon_fitted, Residuals = hon_residuals), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "red") +
  geom_hline(yintercept = 0, color = "blue", linetype = "dashed") +
  labs(title = "HON Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals")

# 2. Homoscedasticity (Constant Variance of Residuals)
# Perform the Breusch-Pagan test for homoscedasticity
ncvTest(aapl_regression)  # For AAPL
ncvTest(hon_regression)   # For HON

# 3. Normality of Residuals
# Normal Q-Q Plot for AAPL
qqPlot(aapl_residuals, main = "AAPL Normal Q-Q Plot", col = "blue")

# Normal Q-Q Plot for HON
qqPlot(hon_residuals, main = "HON Normal Q-Q Plot", col = "red")

# Chi-squared Test for Normality (Shapiro-Wilk Test)
shapiro.test(aapl_residuals)  # For AAPL
shapiro.test(hon_residuals)   # For HON

# Install the nortest package (if not already installed)
install.packages("nortest")

# Load the nortest package
library(nortest)

# Perform Chi-squared test for AAPL residuals
aapl_chisq_test <- pearson.test(aapl_residuals)

# Perform Chi-squared test for HON residuals
hon_chisq_test <- pearson.test(hon_residuals)

# Display the test results
aapl_chisq_test
hon_chisq_test

```

-   The residuals for HON and AAPL are independent.
-   Homoscedasticity (constant variance) is present in the residuals for both AAPL and HON.
-   The residuals are normally distributed by plotting a Normal probability plot of the residuals
-   The residuals are not normally distributed by performing a Chi-squared test for Normality of the residuals

# Portfolio Allocation

```{r}
# Portfolio allocation based on analysis
P <- 40  # 40% allocated to AAPL
Q <- 60  # 60% allocated to HON

# Print the allocation
cat("Portfolio Allocation:\n")
cat("AAPL (Apple Inc.):", P, "%\n")
cat("HON (Honeywell Inc.):", Q, "%\n")
```
